\chapter{Context}

\epigraph{You might have to think  of \\ how you got started \\ sitting in your little room}{Little Room \\ The White Stripes, 2001}

Since joining MIT in summer 2019 I didn't leave the USA, and it's the longest stretch I have had of not visiting my home country Chile. This thesis in particular was written between November 2020 and August 2021, mostly in Boston MA USA, while on a F-1 visa.

\section{Language}

My native language is Spanish, and this thesis was written in English, using the metric system, and the Gregorian year-month-day format.

I tried to avoid violent language, including some widespread conventions on computer science which I hope become obsolete, such as executing a file, instead of running a file.

\section{Software}

This thesis document was written using LaTeX and the Microsoft Visual Studio Code editor, and then exported to the PDF format.

The TinyTrainable library was written in C++, and packaged as an Arduino library, relying on open source library dependencies by Adafruit, Arduino, and Google.

The auxiliary code is a mix of Python scripts, Jupyter Python notebooks, and shell scripts.

The documentation was written using Markdown.

The workshops were taught using the videoconferencing software Zoom, and organized via Google Forms.

\section{Hardware}

This project was written on a 2017 Macbook Air 13-inch, running the macOS 10.15.7 Catalina operating system.

The software library and the software examples were written to be deployed on the Arduino Nano 33\acrshort{BLE}Sense microcontroller.

\section{Collaborators}

Priscilla Capistrano is the senior administrative assistant at the Opera of the Future research group and made sure that everything worked.

Peter Tone is a MIT undergraduate student, who was a researcher, designer, programmer, and tester of the TinyTrainable Arduino library, as part of the MIT UROP program.

Maxwell Wang is a MIT undergraduate student who was a documentation writer, and hardware and software tester, as part of the MIT UROP program.

Roy Macdonald solved my most difficult programming questions, and helped to implement the solutions.

The Council for the Arts at MIT provided the generous funding of the workshop materials.

Renata Gaui designed and created the flyers for the workshops.

Bernardita Moraga packaged and distributed the materials for the workshop in Chile.

\newpage
