\chapter{Conclusions and future work}

In this thesis I have presented all the stages of the design and development of new standalone multimedia instruments using machine learning and microcontrollers, emphasizing \acrshort{AI} ethics. The project includes software examples, hardware suggestions, educational material, and strategies for ethical off-cloud machine learning and creation of custom artisanal databases.

This thesis is also the basis for further research, including the creation of subsequent multimedia instruments and software libraries, the writing of new courses and educational units at the intersection of arts, physical computing, interaction design, and computational ethics.

\section{Contributions}

Concrete contributions:

\begin{enumerate}
  \item Publishing TinyTrainable, a software library for creating instruments with  \acrshort{ML} and microcontrollers for multimedia art.
  \item Designing, writing, and teaching a 4-hour workshop for beginners, enthusiasts, and artists to teach with the TinyTrainable software library.
  \item Publishing code and tutorials for creating custom databases for gesture and speech recognition, and for deleting metadata for privacy concerns.
  \item Publishing custom-trained \acrshort{ML} models for gesture and speech recognition.
  \item Publishing code and tutorials for training \acrshort{ML} algorithms on the cloud and on personal computers for privacy and agency.
  \item Publishing other related software libraries, such as MaquinitasParams for communication with other instruments, and MaquinitasRitmos for rhythmic data.
\end{enumerate}

Abstract contributions:

\begin{enumerate}
 \item Demonstrating how a broader range of people can use \acrshort{ML} to support their creative expression.
 \item Developing strategies for artistic reappropation of the inherent imperfections and shortcomings of \acrshort{ML}.
 \item Sharing the step-by-step process of conceptualizing and implementing a software library for artists.
\end{enumerate}

\section{Lessons learned}

\begin{enumerate}
  \item Writing software for artists is hard.
  \item Writing software libraries for other artists is even harder.
  \item Collaborating with other people is essential to writing usable code and clear educational material.
  \item Documenting all design decisions and steps is key to explaining why and how your project works.
  \item Navigating laws, licenses, and copyright is really hard.
\end{enumerate}

\section{Future work}

\subsection{Hardware for new instruments}

This thesis relies on an Arduino microcontroller because of their open source nature, commercial availability, software and community support, and detailed documentation.

In particular I picked the Arduino Nano 33 \acrshort{BLE} Sense, because of two main reasons at the time this project started in late 2020: it is currently the only Arduino supported by Google's TensorFlow Lite for microcontrollers library and the HarvardX certificate on tiny \acrshort{ML}. Also, because of its convenience of having embedded sensors, which makes it simpler and cheaper to acquire data for live interaction and for building custom databases, eliminating barriers to instrument makers and prototypers.

Microcontrollers come and go. Most probably this Arduino will be discontinued, but the strategies and software presented in this thesis can be adapted to other microcontrollers and software architectures. I am particularly looking forward to this thesis project being ported to other Arduinos, PJRC Teensy, and Adafruit Circuit Playground. They feature different software stacks, such as Python instead of C++, and through that cater to a wider community of people building multimedia instruments and artwork.

In terms of the outputs of the Tiny Trainable Instruments, I focused on creating many parallel multimedia approaches, including making sounds with piezo buzzers and MIDI, manipulating light with LEDs, creating movement and rhythm with servo motors, and printing text with thermal printers and screens. This is to appeal to a large audience of artists and learners, interested in different mediums. I hope this thesis project inspires the addition of further outputs, and that people can contribute back to the library to share these new capabilities with everyone.

The Tiny Trainable Instruments are built with prototyping electronic breadboards, to make explicit their open-endedness, and to promote experimentation and lower barriers. A further iteration of this project that I wish to make is creating custom \acrshortpl{PCB} with fixed wiring, and also enclosures and packaging.

\subsection{Software for new instruments}

This thesis has been published as an open source software library for Arduino. It promotes modularity and adaptability: a Tiny Trainable Instrument can be any combination of the multiple inputs and outputs. The file structure of the source code and the software dependencies of this library were also written with flexibility in mind, to encourage the remix and adaptation of this library to further projects.

A challenging aspect of this project has been the breadth of the disciplines combined, and its novel application of \acrshort{ML} in microcontrollers. There is a community of makers creating open source standalone multimedia instruments, but the required skills are still hard to acquire. I hope this project helps foster a new generation of instrument makers, who can learn from this project, like I have learned from my favorite makers.

Additionally, the principles of this project, including being as cheap as possible and as open as possible, are designed to encourage experimentation and hacking, but also can pose additional challenges. I hope this project encourages people to learn how to make instruments, and also engages in discourse about the creation of new curricula for the next generation of instrument makers and artists.

Another challenging aspect of writing software for multimedia instruments is its licensing, both choosing a license and also respecting and understanding the license of other code and resources we are using. The dependencies of this software are mostly other libraries by Arduino, Google, Adafruit, and with different licenses including public domain, MIT, and Apache. I hope this document helps to navigate these legal complexities and that this project helps artists and enthusiasts to navigate this landscape and overcome these barriers.

\subsection{Educational impact}

This project was built to inspire and celebrate a new generation of coursework, workshops, and books, in the disciplines of ethical \acrshort{ML} and microcontroller-based instruments.

I hope that this thesis project is adopted by educators, to introduce students to \acrshort{ML}, physical computing, media arts, and computational ethics. It would be amazing if aspects of Tiny Trainable Instruments could be incorporated into new and existing music, arts, sculpture, and computer science curricula, to create a new wave of instrument makers and media artists.

Since the \acrshort{ML} resources we are using are very limited in comparison to state-of-the-art industrial systems, the Tiny Trainable Instruments are slower and more prone to errors than their commercial counterparts. This is an opportunity for creating new artistic experiences that exploit the pitfalls of \acrshort{ML}, and in turn help advance the understanding and critique of these emerging computational techniques.
