\chapter{Tiny trainable instruments}

\epigraph{All the modern things \\ have always existed \\ they've just been waiting}{The Modern Things \\ Björk, 1995}

This chapter describes what this project is, its design principles, and scope of this thesis.

\section{Design principles}

\begin{enumerate}
  \item Affordable
  \item Hackable
  \item Open
  \item Private
\end{enumerate}

\subsection{Cheap}

The materials for this thesis 

\section{Antidisciplinarity}

This project draws from different disciplines, all intertwined and in different capacities, including:

\begin{enumerate}
  \item Artificial intelligence
  \item Electrical engineering
  \item Computer science
  \item Media arts
  \item Music
\end{enumerate}

\section{Technology}

This project is built with microcontrollers and 

This project is based on the Arduino Nano 33 BLE Sense microcontroller, and its software dependencies include the Arduino KNN library, the Arduino TensorFlow Lite Micro library, and Adafruit software libraries. 

\section{Open}

All examples included with this library were written with the aim of showing the fundamentals of how to build the instruments and different machine learning enabled manipulation of multimedia material, so that people could build on top of it and make it their own, by changing the values of variables and adding more functionalities.

\section{Philosophy and experience}

Throughout this project, the magic number was 3. The machine learning algorithms were hardcoded to be able to distinguish between 3 different categories: 3 colors, 3 physical gestures, 3 sound utterances.

\section{Inputs}

We are using the RGB color, proximity, gyroscope, accelerometer, and microphone sensors on the microcontroller, in order to capture the Inputs

\begin{enumerate}
  \item Color
  \item Gesture
  \item Speech
\end{enumerate}

\subsection{Color}

This approach uses the RGB color sensor from the microcontroller, with the auxiliary help from the proximity sensor, that is used to capture color information at a certain distance threshold.

The data is passed to a k-Nearest-Neighbor algorithm, programmed using the Arduino KNN library.

\subsection{Gesture}

This input uses the information from the Inertial Measurement Unit (IMU) of the microcontroller, including a gyroscope and accelerometer. It captures data after a certain threshold of movement is detected.

The data is passed to a TensorFlow neural network, programmed using the Arduino TensorFlow Lite library, and based on the included magic$\_$wand example.

\subsection{Speech}

This input uses the information from the microphone of the microcontroller.

The data is passed to a TensorFlow neural network, programmed using the Arduino TensorFlow Lite library, and based on the included micro$\_$speech example.

\section{Outputs}

The different outputs were picked, because of their low cost, ubiquity, and possibilities of expansion and combining them.

\subsection{Buzzer}

This output creates pitched sound, by using a PWM output.

\subsection{Servo motor}

This output creates movement and through that, rhythmic sounds.

The main inspiration for this output was the emerging use of motor-activated percussive instruments, such as the Polyend Perc.

\subsection{MIDI}

We wrote functionalities to manipulate MIDI instruments, and included examples to interface with some popular and cheap MIDI instruments, such as the Korg volca beats.

We included examples for rhythmic and melodic elements, using two very ubiquitous and inexpensive MIDI musical instruments, which are the Korg volca beats, and the Korg volca keys.

\subsection{Thermal printer}

A thermal printer is the basis for creating written and literary output, inspired by the field of computational poetry.

We used the popular Adafruit Thermal printer kit, which is documented on their website and includes a software library, distributed over GitHub and Arduino IDE, and also as a submodule on this project's TinyTrainable software library.

\section{Development}

This thesis has been developed with the invaluable help of undergrad researchers Peter Tone and Maxwell Wang.

They have cloned both repositories, the main one and the Arduino library one, and have continuously submitted pull requests with their contributions.

Peter Tone has helped with research in data structures, library writing, and we have shared back and forth code, going from experimental proofs of concepts, and has also helped with the design of the user-facing library.

Maxwell Wang has proofread the code, has run the examples, and has helped with the writing of the documentation for self-learners and for the workshops.

We all share a Google Drive folder, where we all share notes about our research and development of the library and the educational material.

\section{Code}

This thesis is distributed as a repository, hosted on the GitHub platform, and available at https://github.com/montoyamoraga/tiny-trainable-instruments.

The auxiliary files, such as the LaTeX project for this document, and the auxiliary Jupyter notebooks, and documentation and tutorials are included on this repository.

The main software component of this project is the TinyTrainable library, available at https://github.com/montoyamoraga/TinyTrainable and also through the Arduino IDE.

The code included on this library is distributed on the folders:

\begin{enumerate}
  \item examples/
  \item src/
\end{enumerate}

\subsection{src/}

The source code for where there is a TinyTrainable.h and TinyTrainable.cpp file where we included all the basic functionality of the library. Additional subfolders include

\subsubsection{inputs/}

Base class Input and inherited classes for each one of the other inputs.

\subsubsection{outputs/}

Base class Output and inherited classes for each one of the other outputs.

\subsubsection{tensorflow/}

Auxiliary files, copied from the examples from the Arduino TensorFlow Lite that we are building on top of, and also from the newer TinyML library by the EdX team. These, unless otherwise noted, are included without modifications and distributed through the Apache License included on each file's headers.

\section{Opera of the Future projects}

During the development of this thesis, I have been fortunate to collaborate on different capacities with other thesis projects by classmates at Opera of the Future, which has directly inspired my work.

\subsection{Squishies, by Hannah Lienhard}

Squishies is Hannah Lienhard's master's thesis, and consists of novel squishable interfaces for musical expression. We shared discussions about low-level sound design, code reusability, sound art education, and digital instruments. We were part of a master's thesis working group, facilitated by Roya Moussapour with two other Media Lab classmates, where we workshopped drafts of our thesis. This practice and feedback has been critical in shaping the language and discourse of this thesis document.

\subsection{Fluid Music, by Charles Holbrow}

Fluid Music is Charles Holbrow's PhD thesis. It is a library for library design, documentation for contributors. The design of the interface, documentation, and scope of the thesis were a direct influence on the documentation and API coding style of this project.

Overall thoughts: I think it would be great if your Chapter 3 clearly led you to the design principles you lay out in Chapter 4. So, why is “open” important -> because of the collaborations or arduino libraries you used. Why is “cheap”/”privacy” -> coded bias and ability to tinker, etc.
